\bfseries\LARGE\singlespacing{Transmission of airborne respiratory infections with and without air cleaners in a Swiss school under non-pandemic conditions in 2023: A modeling study of epidemiological, environmental, and molecular data}

### 2023-07-05 doc/paper/manuscript.tex
Replace pathogen with virus

\noindent\textbf{Methods and Findings:} We collected epidemiological (absences related to respiratory infections), environmental (CO$_2$ and particle concentrations), and molecular data (bioaerosol and saliva samples) over seven weeks from 16 January to 11 March 2023 in a secondary school (n=38 students in two classes) in Switzerland. We performed laboratory and molecular analyses to detect a panel of 19 respiratory viruses. We installed air cleaners in a cross-over design and compared positive samples and particle concentrations between weeks with and without air filters. We modeled the number of new infections using a Bayesian negative binomial regression model, adjusting for the incubation period, the number of recent and cumulative infections, the ventilation rate, the number of students in class, and community transmission. \\
Molecular analysis detected multiple viruses in saliva samples throughout the study (50/448 positive: 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza virus), but rarely in bioaerosol samples (2/105 positive: 1~rhinovirus and 1~adenovirus) and on the filters of air cleaners (4/160~positive: 1~influenza~B, 1~HRV, 1~adenovirus, and 1~SARS-CoV-2). The detection rate of positive saliva samples was comparable between study conditions (adjusted risk ratio 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61). Daily average CO$_2$ levels were 1,769$\pm$391\,ppm ($\pm$ standard deviation) with air cleaners and 1,636$\pm$341\,ppm without air cleaners, corresponding to similar air change rates (0.74$\pm$0.41\,h$^{-1}$ vs. 0.68$\pm$0.32\,h$^{-1}$). Daily average aerosol number concentrations were 27$\pm$34\,1/cm$^3$ with and 95$\pm$81\,1/cm$^3$ without air cleaners, corresponding to an estimated adjusted decrease of 77\% (95\%-CrI 63\% to 86\%). Absences related to respiratory infections were less frequent with air cleaners in place (13~absences with vs. 21~absences without air cleaners). The estimated risk of infection tended to be lower with air cleaners (adjusted risk ratio 0.72, 95\%-Credible Interval 0.44 to 1.15). \\
% Study limitations include unknown specific causes of absences related to respiratory infections (i.e which pathogen caused the sickness). Furthermore, students occasionally had lessons outside the classroom, thus limiting the potential effectiveness of air cleaners.
\medskip %\vspace{-1.3em}

@@

The transmission of respiratory infections such as SARS-CoV-2 and influenza presents significant challenges in terms of mitigation and control. Person-to-person transmission primarily occurs through the release of respiratory particles containing the respective pathogens. The recent focus has been on small respiratory particles ($\leq$ 5$\mu$m) called aerosols, which have been found to carry the majority of viruses during respiratory activities\cite{Fennelly2020}. Unlike larger respiratory droplets that tend to settle quickly, aerosols can remain suspended in the air for several hours and their respiratory plumes can reach more than 2m\cite{Coleman2022,Wang2020,Heneghan2021}, thus exceeding the recommended rules for social distancing by public health organizations\cite{Bazant2021PNAS,Trivedi2021PhyFluids,Lindsley2010CID}.

The generation, spread, and concentration of bioareosols can be influenced with infection control measures. For example, compulsory face mask wearing reduced the transmission of SARS-CoV-2 during the COVID-19 pandemic\cite{Gettings2021,Heinsohn2022,Leech2022PNAS,Mitze2020PNAS}. More generally, improved ventilation systems are considered crucial for healthy indoor environments both during pandemic and non-pandemic times\cite{Wang2021,Morawska2021}. Portable air-filtration devices (air cleaners) may present another cost-effective alternative to upgrading ventilation systems, but the effects of air cleaners on the transmission of respiratory infections are less clear. Various studies showed that air cleaners reduce the concentration of aerosols and particle matter in indoor air\cite{Park2020Build,Buising2022InfContr,Ren2021Dent,Banholzer2023PLoSMed}. These findings suggest potential protection from respiratory infection as long-term exposure to higher aerosol and particle matter mass concentration have been associated with increased risks of respiratory infections\cite{Gordon2014Resp} and, more recently, with the incidence and mortality of COVID-19\cite{Kelly2023Atmos,Veronesi2022OccEnvMed,Travaglio2021EnvPoll,DeAngelis2021EnvRes}. Furthermore, a recent study showed that air cleaners effectively removed SARS-CoV-2 bioaerosols in a hospital ward\cite{Morris2022}. As a result, studies have simulated the potential of air cleaners in reducing indoor transmission risks of SARS-CoV-2\cite{Lindsley2021} and other respiratory pathogens\cite{Cortellessa2023Build}. However, empirical evidence linking the use of air cleaners with a reduced risk of infection is lacking. Thus, it remains unclear whether the reduction in particle concentrations and the removal of bioaerosols from the air translates into lower transmission of respiratory infections.

@@

A detailed description of the model and choice of priors for all model parameters are provided in Text~\zref{sec:transmission-model} in \supp. Importantly, since multiple pathogens were prevalent over the study period, we considered the distribution of the incubation period to be a weighted mix of the pathogen-specific distributions. The weights were inferred based on the proportion of positive saliva samples for each pathogen per week. As a result, the distribution of the incubation period used for modeling could change every week depending on the proportion of pathogens that were detected in the students' saliva. Moreover, respiratory cases by date of symptom onset were under-reported for weekends and thus a re-weighting scheme was used to account for weekday effects. \medskip

@@

\noindent Molecular analysis determined which saliva samples were positive for a respiratory pathogen. We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (Text~\zref{sec:multinomial-model} in \supp) where the expected number of positive samples for a pathogen was linked to the presence of air cleaners, adjusting for class-specific effects and the number of susceptibles for each pathogen. The latter was computed as the difference between the number of students per class and the cumulative number of positive samples. As a sensitivity analysis, we lagged the test results by one test date (\ie from Tuesday to last Thursday and from Thursday to Tuesday) and one test week (\ie from Tuesday to last Tuesday and Thursday to last Thursday), to consider the unknown delay when students were infected and tested positive.

@@

We analyzed 448~saliva samples over the study across classes. We detected 15~influenza~B~(Flu B), 15~rhinoviruses~(HRV), 14~adenovirus (AdV), 3~SARS-CoV-2~(CoV), 2~metapneumoviruses~(MPV), and 1~parainfluenza~virus~(PIV) (\Cref{fig:molecular-descriptives}a). There was an equal number of positive samples with and without air cleaners (\Cref{fig:molecular-descriptives}b). Taking into account the number of samples and the number of suceptible students, the positivity rate remained comparable between study conditions (adjusted risk ratio 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61), and was not sensitivity to infection to testing delays (Figure~\zref{fig:mol-estimation-results-sensitivity} in \supp). The distribution of positive samples varied between classes (\Cref{fig:molecular-descriptives}c). For example, in the first three study weeks all except for one sample were positive for AdV in class~A, while the vast majority of positive Flu~B samples were found in class~B. The pathogen-specific variations between classes are suggestive of transmission occurring within classes.

@@

\begin{figure}[!htpb]
    \centering
    \includegraphics{../../results/mol-data/descriptives.pdf}
    \caption{\textbf{Analysis of molecular data}. \textbf{(a)}~Number of positive saliva samples by pathogen. \textbf{(b)}~Number of positive saliva samples by study condition. \textbf{(c)}~Number of positive saliva samples by study week and class. Pathogens: Influenza B (Flu B), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza virus (PIV).}
    \label{fig:molecular-descriptives}
\end{figure}

@@

In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners in reducing the transmission of SARS-CoV-2 in two Swiss schools. We only found a detectable effect on the risk of infection for mask wearing, but not air cleaners. We argued that air cleaners may have had less potential to reduce infections because they were implemented at the end of the study when most students had already been infected with SARS-CoV-2. In this study, we used a cross-over design to estimate the effectiveness of air cleaners. We found similar reductions in particle concentrations and a tendency that the relative risk of infection was lower with air cleaners. To some extent, the differences between the two studies may be explained by different adjustments. While we could not adjust for ventilation in our previous study, the estimates of this study were adjusted for the daily air change rate, which correlated with the relative risk of infection (see Text~\zref{sec:detailed-molecular} in \supp). Furthermore, both studies were conducted during different times and the distribution of respiratory pathogens varied considerably. During the COVID-19 pandemic (previous study), we could almost only detect SARS-CoV-2, whereas a range of respiratory pathogens was detected during non-pandemic times (this study).

@@

Our study indicated that air cleaners can reduce the risk of infection, although we we only were able tho show a trend. Other infection control measures, such as universal mask use, appear to be more effective (REFS). There may be several reasons for this. Compared to \eg mask wearing, air cleaners cannot influence transmission outside the classrooms. Furthermore, air cleaners are unable to prevent close range high particle density aerosol transmission. By contrast, face masks can also reduce the generation of airborne infectious particles and can thus mitigate transmission over short ranges. Evidence from our previous\cite{Banholzer2023PLoSMed} and this study further suggest that air cleaners are more effective in the removal of the larger particles ($>5\mu$m) than the smaller particles ($\leq5\mu$m). Yet, a larger proportion of respiratory pathogens are carried in the smaller particles\cite{Fennelly2020}.

@@

We detected few respiratory pathogens in bioaerosol samples (3~positive bioaoerosl and 50~positive saliva samples), especially when comparing to our previous study\cite{Banholzer2023PLoSMed} (10~positive bioaerosol and 21~positive saliva samples). The number of pathogens detected on the filters of the air cleaners was also lower (this study: 4~positive samples in 160~swabs, previous study: 8~positive samples in 80~swabs). The bioaerosol sampling devices were the same in both studies and technical issues were not observed. It is also unlikely that environmental factors are responsible for the differences. Temperature and humidity were similar, and CO$_2$ levels were even higher in this study (1,702$\pm$370\,ppm in this vs 1,064$\pm$232\,ppm in the previous study), suggesting that classrooms were not as well ventilated as during the COVID-19 pandemic. Therefore, we think that the differences in the number of positive bioaerosols samples between both studies is most likely related to differences in the distribution of respiratory pathogens. Strikingly, we detected a wide range of respiratory pathogens (Flu B, HRV, AdV, etc) in this study, as compared to only SARS-CoV-2 in our previous study. The shift in the pattern of respiratory viruses has been observed in other studies (xxx). SARS-CoV-2 may be present in higher concentrations in air and may be more easily detected in bioaerosol samples due to differences in physicochemical properties of virus-laden aerosols (\eg physical size, viral load, and infectivity) that influence the generation and survival of airborne viral RNA\cite{Wang2021,Sattar2016Book}. In addition, although respiratory pathogens have been detected in bioaerosols (\eg Flu\cite{Bischoff2013JID,Pan2017mSphere}, HRV\cite{Myatt2004AJRCCM,Fabian2011JAMPDD}, AdV\cite{Nguyen2017OFID,Pan2017mSphere}, and CoV\cite{Guo2020EID,Liu2020Nature}), bioaerosol sampling remains a technical challenge\cite{Wang2021 ,Belser2023PLOSPath,Bekking2019IORV,Mainelis2020AST}.

@@

The low rate of positive bioaerosol samples also means that it is unclear to what extent students airborne transmission occurred in classrooms. It may be possible that students had rather little exposure to respiratory pathogens in schools and acquired their infections elsewhere. However, the distribution of saliva samples between classes shows remarkable differences. AdV was spreading entirely in class~A in the first three study weeks and only two infections with Flu~B occurred over the study. By contrast, Flu~B was spreading throughout the study in class~B and infections with AdV were only detected in the last study week. These class-specific patterns in the spread of respiratory pathogens indicate that transmission of respiratory infections may have occurred at least to some extent within the classrooms.

@@

Our study has several limitations. First, aerosol measurements and molecular detection of pathogens in bioaerosol samples document exposure, but not necessarily transmission and the direction of transmission (human to air, air to human) cannot be determined. Nevertheless, the pathogen-specific distribution of positive saliva samples between classes indicate potential transmission within classrooms. Second, it was not possible to ascertain the specific causes for the absences based on the epidemiological data. Therefore, some absences may have falsely been attributed to respiratory infections and we could only approximate the incubation period for each epidemiological case. Third, the effectiveness of air cleaners was somewhat limited because students occasionally had lessons outside the classroom. This is in contrast to other infection control measures such as face masks that had to be used in every indoor space during pandemic times.

@@

In conclusion, using epidemiological, environmental, and molecular data, our study shows that considerable transmission of respiratory pathogens other than SARS-CoV-2 occurred in school rooms under non-pandemic conditions. Air cleaners significantly reduced aerosol concentrations, but airborne respiratory viruses were rarely detected and if only viruses other than SARS-CoV-2. The epidemiologic transmission model estimated a trend toward a reduction in infections with air filters. Future studies may use our multiple-measurement approach to assess whether professional building ventilation, which should be preferred over air cleaners in the long term, and other infection control measures reduce the risk of airborne transmission. Whether their effectiveness varies with the seasonal distribution of respiratory pathogens under non-pandemic conditions remains to be investigated.

>>>>

\noindent\textbf{Methods and Findings:} We collected epidemiological (absences related to respiratory infections), environmental (CO$_2$ and particle concentrations), and molecular data (bioaerosol and saliva samples) over seven weeks from 16 January to 11 March 2023 in a secondary school (n=38 students in two classes) in Switzerland. We performed laboratory and molecular analyses to detect a panel of 19 respiratory viruses. We installed air cleaners in a cross-over design and compared positive samples and particle concentrations between weeks with and without air filters. We modeled the number of new infections using a Bayesian negative binomial regression model, adjusting for the incubation period, the number of recent and cumulative infections, the ventilation rate, the number of students in class, and community transmission. \\
Molecular analysis detected multiple viruses in saliva samples throughout the study (50/448 positive: 15~influenza B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, and 1~parainfluenza virus), but rarely in bioaerosol samples (2/105 positive: 1~rhinovirus and 1~adenovirus) and on the filters of air cleaners (4/160~positive: 1~influenza~B, 1~HRV, 1~adenovirus, and 1~SARS-CoV-2). The detection rate of positive saliva samples was comparable between study conditions (adjusted risk ratio 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61). Daily average CO$_2$ levels were 1,769$\pm$391\,ppm ($\pm$ standard deviation) with air cleaners and 1,636$\pm$341\,ppm without air cleaners, corresponding to similar air change rates (0.74$\pm$0.41\,h$^{-1}$ vs. 0.68$\pm$0.32\,h$^{-1}$). Daily average aerosol number concentrations were 27$\pm$34\,1/cm$^3$ with and 95$\pm$81\,1/cm$^3$ without air cleaners, corresponding to an estimated adjusted decrease of 77\% (95\%-CrI 63\% to 86\%). Absences related to respiratory infections were less frequent with air cleaners in place (13~absences with vs. 21~absences without air cleaners). The estimated risk of infection tended to be lower with air cleaners (adjusted risk ratio 0.72, 95\%-Credible Interval 0.44 to 1.15). \\
% Study limitations include unknown specific causes of absences related to respiratory infections (i.e which virus caused the sickness). Furthermore, students occasionally had lessons outside the classroom, thus limiting the potential effectiveness of air cleaners.
\medskip %\vspace{-1.3em}

@@

The transmission of respiratory infections such as SARS-CoV-2 and influenza presents significant challenges in terms of mitigation and control. Person-to-person transmission primarily occurs through the release of respiratory particles containing the respective viruses. The recent focus has been on small respiratory particles ($\leq$ 5$\mu$m) called aerosols, which have been found to carry the majority of viruses during respiratory activities\cite{Fennelly2020}. Unlike larger respiratory droplets that tend to settle quickly, aerosols can remain suspended in the air for several hours and their respiratory plumes can reach more than 2m\cite{Coleman2022,Wang2020,Heneghan2021}, thus exceeding the recommended rules for social distancing by public health organizations\cite{Bazant2021PNAS,Trivedi2021PhyFluids,Lindsley2010CID}.

The generation, spread, and concentration of bioareosols can be influenced with infection control measures. For example, compulsory face mask wearing reduced the transmission of SARS-CoV-2 during the COVID-19 pandemic\cite{Gettings2021,Heinsohn2022,Leech2022PNAS,Mitze2020PNAS}. More generally, improved ventilation systems are considered crucial for healthy indoor environments both during pandemic and non-pandemic times\cite{Wang2021,Morawska2021}. Portable air-filtration devices (air cleaners) may present another cost-effective alternative to upgrading ventilation systems, but the effects of air cleaners on the transmission of respiratory infections are less clear. Various studies showed that air cleaners reduce the concentration of aerosols and particle matter in indoor air\cite{Park2020Build,Buising2022InfContr,Ren2021Dent,Banholzer2023PLoSMed}. These findings suggest potential protection from respiratory infection as long-term exposure to higher aerosol and particle matter mass concentration have been associated with increased risks of respiratory infections\cite{Gordon2014Resp} and, more recently, with the incidence and mortality of COVID-19\cite{Kelly2023Atmos,Veronesi2022OccEnvMed,Travaglio2021EnvPoll,DeAngelis2021EnvRes}. Furthermore, a recent study showed that air cleaners effectively removed SARS-CoV-2 bioaerosols in a hospital ward\cite{Morris2022}. As a result, studies have simulated the potential of air cleaners in reducing indoor transmission risks of SARS-CoV-2\cite{Lindsley2021} and other respiratory viruses\cite{Cortellessa2023Build}. However, empirical evidence linking the use of air cleaners with a reduced risk of infection is lacking. Thus, it remains unclear whether the reduction in particle concentrations and the removal of bioaerosols from the air translates into lower transmission of respiratory infections.

@@

A detailed description of the model and choice of priors for all model parameters are provided in Text~\zref{sec:transmission-model} in \supp. Importantly, since multiple viruses were prevalent over the study period, we considered the distribution of the incubation period to be a weighted mix of the virus-specific distributions. The weights were inferred based on the proportion of positive saliva samples for each virus per week. As a result, the distribution of the incubation period used for modeling could change every week depending on the proportion of viruses that were detected in the students' saliva. Moreover, respiratory cases by date of symptom onset were under-reported for weekends and thus a re-weighting scheme was used to account for weekday effects. \medskip

@@

\noindent Molecular analysis determined which saliva samples were positive for a respiratory virus. We analyzed the number of positive saliva samples with a Bayesian Multinomial logistic regression model (Text~\zref{sec:multinomial-model} in \supp) where the expected number of positive samples for a virus was linked to the presence of air cleaners, adjusting for class-specific effects and the number of susceptibles for each virus. The latter was computed as the difference between the number of students per class and the cumulative number of positive samples. As a sensitivity analysis, we lagged the test results by one test date (\ie from Tuesday to last Thursday and from Thursday to Tuesday) and one test week (\ie from Tuesday to last Tuesday and Thursday to last Thursday), to consider the unknown delay when students were infected and tested positive.

@@

We analyzed 448~saliva samples over the study across classes. We detected 15~influenza~B~(Flu B), 15~rhinoviruses~(HRV), 14~adenovirus (AdV), 3~SARS-CoV-2~(CoV), 2~metapneumoviruses~(MPV), and 1~parainfluenza~virus~(PIV) (\Cref{fig:molecular-descriptives}a). There was an equal number of positive samples with and without air cleaners (\Cref{fig:molecular-descriptives}b). Taking into account the number of samples and the number of suceptible students, the positivity rate remained comparable between study conditions (adjusted risk ratio 0.93, 95\%-Credible interval [CrI] 0.49 to 1.61), and was not sensitivity to infection to testing delays (Figure~\zref{fig:mol-estimation-results-sensitivity} in \supp). The distribution of positive samples varied between classes (\Cref{fig:molecular-descriptives}c). For example, in the first three study weeks all except for one sample were positive for AdV in class~A, while the vast majority of positive Flu~B samples were found in class~B. The virus-specific variations between classes are suggestive of transmission occurring within classes.

@@

\begin{figure}[!htpb]
    \centering
    \includegraphics{../../results/mol-data/descriptives.pdf}
    \caption{\textbf{Analysis of molecular data}. \textbf{(a)}~Number of positive saliva samples by virus. \textbf{(b)}~Number of positive saliva samples by study condition. \textbf{(c)}~Number of positive saliva samples by study week and class. Viruses: Influenza B (Flu B), rhinovirus (HRV), adenovirus (AdV), SARS-CoV-2 (CoV), metapneumovirus (MPV), parainfluenza virus (PIV).}
    \label{fig:molecular-descriptives}
\end{figure}

@@

In a previous study\cite{Banholzer2023PLoSMed}, we estimated the effectiveness of mask wearing and air cleaners in reducing the transmission of SARS-CoV-2 in two Swiss schools. We only found a detectable effect on the risk of infection for mask wearing, but not air cleaners. We argued that air cleaners may have had less potential to reduce infections because they were implemented at the end of the study when most students had already been infected with SARS-CoV-2. In this study, we used a cross-over design to estimate the effectiveness of air cleaners. We found similar reductions in particle concentrations and a tendency that the relative risk of infection was lower with air cleaners. To some extent, the differences between the two studies may be explained by different adjustments. While we could not adjust for ventilation in our previous study, the estimates of this study were adjusted for the daily air change rate, which correlated with the relative risk of infection (see Text~\zref{sec:detailed-molecular} in \supp). Furthermore, both studies were conducted during different times and the distribution of respiratory viruses varied considerably. During the COVID-19 pandemic (previous study), we could almost only detect SARS-CoV-2, whereas a range of respiratory viruses was detected during non-pandemic times (this study).

@@

Our study indicated that air cleaners can reduce the risk of infection, although we we only were able tho show a trend. Other infection control measures, such as universal mask use, appear to be more effective (REFS). There may be several reasons for this. Compared to \eg mask wearing, air cleaners cannot influence transmission outside the classrooms. Furthermore, air cleaners are unable to prevent close range high particle density aerosol transmission. By contrast, face masks can also reduce the generation of airborne infectious particles and can thus mitigate transmission over short ranges. Evidence from our previous\cite{Banholzer2023PLoSMed} and this study further suggest that air cleaners are more effective in the removal of the larger particles ($>5\mu$m) than the smaller particles ($\leq5\mu$m). Yet, a larger proportion of respiratory viruses are carried in the smaller particles\cite{Fennelly2020}.

@@

We detected few respiratory viruses in bioaerosol samples (3~positive bioaoerosl and 50~positive saliva samples), especially when comparing to our previous study\cite{Banholzer2023PLoSMed} (10~positive bioaerosol and 21~positive saliva samples). The number of viruses detected on the filters of the air cleaners was also lower (this study: 4~positive samples in 160~swabs, previous study: 8~positive samples in 80~swabs). The bioaerosol sampling devices were the same in both studies and technical issues were not observed. It is also unlikely that environmental factors are responsible for the differences. Temperature and humidity were similar, and CO$_2$ levels were even higher in this study (1,702$\pm$370\,ppm in this vs 1,064$\pm$232\,ppm in the previous study), suggesting that classrooms were not as well ventilated as during the COVID-19 pandemic. Therefore, we think that the differences in the number of positive bioaerosols samples between both studies is most likely related to differences in the distribution of respiratory viruses. Strikingly, we detected a wide range of respiratory viruses (Flu B, HRV, AdV, etc) in this study, as compared to only SARS-CoV-2 in our previous study. The shift in the pattern of respiratory viruses has been observed in other studies (xxx). SARS-CoV-2 may be present in higher concentrations in air and may be more easily detected in bioaerosol samples due to differences in physicochemical properties of virus-laden aerosols (\eg physical size, viral load, and infectivity) that influence the generation and survival of airborne viral RNA\cite{Wang2021,Sattar2016Book}. In addition, although respiratory viruses have been detected in bioaerosols (\eg Flu\cite{Bischoff2013JID,Pan2017mSphere}, HRV\cite{Myatt2004AJRCCM,Fabian2011JAMPDD}, AdV\cite{Nguyen2017OFID,Pan2017mSphere}, and CoV\cite{Guo2020EID,Liu2020Nature}), bioaerosol sampling remains a technical challenge\cite{Wang2021 ,Belser2023PLOSPath,Bekking2019IORV,Mainelis2020AST}.

@@

The low rate of positive bioaerosol samples also means that it is unclear to what extent students airborne transmission occurred in classrooms. It may be possible that students had rather little exposure to respiratory viruses in schools and acquired their infections elsewhere. However, the distribution of saliva samples between classes shows remarkable differences. AdV was spreading entirely in class~A in the first three study weeks and only two infections with Flu~B occurred over the study. By contrast, Flu~B was spreading throughout the study in class~B and infections with AdV were only detected in the last study week. These class-specific patterns in the spread of respiratory viruses indicate that transmission of respiratory infections may have occurred at least to some extent within the classrooms.

@@

Our study has several limitations. First, aerosol measurements and molecular detection of viruses in bioaerosol samples document exposure, but not necessarily transmission and the direction of transmission (human to air, air to human) cannot be determined. Nevertheless, the virus-specific distribution of positive saliva samples between classes indicate potential transmission within classrooms. Second, it was not possible to ascertain the specific causes for the absences based on the epidemiological data. Therefore, some absences may have falsely been attributed to respiratory infections and we could only approximate the incubation period for each epidemiological case. Third, the effectiveness of air cleaners was somewhat limited because students occasionally had lessons outside the classroom. This is in contrast to other infection control measures such as face masks that had to be used in every indoor space during pandemic times.

@@

In conclusion, using epidemiological, environmental, and molecular data, our study shows that considerable transmission of respiratory viruses other than SARS-CoV-2 occurred in school rooms under non-pandemic conditions. Air cleaners significantly reduced aerosol concentrations, but airborne respiratory viruses were rarely detected and if only viruses other than SARS-CoV-2. The epidemiologic transmission model estimated a trend toward a reduction in infections with air filters. Future studies may use our multiple-measurement approach to assess whether professional building ventilation, which should be preferred over air cleaners in the long term, and other infection control measures reduce the risk of airborne transmission. Whether their effectiveness varies with the seasonal distribution of respiratory viruses under non-pandemic conditions remains to be investigated.
